% $Id: radar.tex,v 1.5 2007/06/09 18:58:09 jdsahr Exp $

\documentclass[12pt]{book}

\nonstopmode

\usepackage{makeidx}
\usepackage{epsfig}

% \usepackage{layout}

\input{macros}

%\theoremstyle{plainn}
% \newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{defn}{Definitions}[section]

%\newcommand{\answer}[1]{}
\newcommand{\answer}[1]{\noindent \textbf{Answer}: {#1} }

% \newfont{\jds}{cmr12 at 24pt}

\title{Radar Remote Sensing of Deep Fluctuating Targets}
\author{John D. Sahr}

\makeindex
\typeout{hello}
\begin{document}

\maketitle

\rule{0mm}{0mm}
\vfill

\centerline{\copyright{1999--2017} by John D.~Sahr}

\vfill

\rule{0mm}{0mm}

\tableofcontents

\newpage
% \layout

\part{Basics}

\include{chapter1}
\include{chapter2}
\include{chapter3}

\part{Radar Waveforms}

\include{chapter4}
\include{chapter5}

\include{chapter-CorrelationCodes}

\part{Estimation}

\include{chapter-Estimation}
\include{chapter-Detection}

\part{Appendices}

\appendix

\include{chapter-DopplerProp}
\include{chapter-Math-Reference}
\include{chapter-Ambiguity}
\include{chapter-CramerRao}
\include{chapter-digitizers}
\include{chapter-analytic}

\chapter{Hysell--Chau}
HC6 refers to the paper by Hysell and Chau, 2006\cite{hysell2006optimal}.
visibility from a baseline pair
\begin{eqnarray}
V(k(\vec{d}_n - \vec{d}_m)) &=& \int_{4\pi} A_n(\Omega) A^\ast_m(\Omega) B(\Omega) \exp(j k (\vec{d}_n - \vec{d}_m)\cdot \vec{\Omega}) \; d\Omega \\		
V(k\vec{d}_{nm}) &=& \int_{4\pi} B'_{nm}(\Omega) \exp(j k \vec{d}_{nm} \cdot \vec{\Omega}) \; d\Omega \label{e:vis0} \\
A_n(\Omega)      &=& n\textrm{th antenna response in direction } \Omega \\
B(\Omega)           &=& \textrm{image brightness in direction } \Omega \\
B'_{nm}(\Omega) &=& A_n(\Omega)A^\ast_m(\Omega) B(\Omega) \\
\vec{d}_n             &=& n\textrm{th antenna position} \\
\vec{d}_{nm}        &=&  \vec{d}_n - \vec{d}_m \;\;\; \textrm{antenna pair separation; baseline}
\end{eqnarray}
Note that the brightness $B$ properly has units of Watts/sr, so that the integral \eqref{vis0} can be interpreted as a dot 
product between $\exp(j k \vec{d}_{nm} \cdot \vec{\Omega})$ and $B'_{nm}(\Omega) \; d\Omega$ when both are
suitably sampled over the domain of integration.  In this case we may write
\begin{equation}
V(k\vec{d}_{nm}) = (B'(\Omega^i)d\Omega)\exp(j k \vec{d}_{nm} \cdot \vec{\Omega^i})  
\end{equation}
or more simply
\begin{eqnarray}
V(k\vec{d}_{nm}) &=& B^i h^i_{nm} \\
B^i       &=& B'(\Omega^i) d\Omega \\
h^i_{nm} &=&\exp(j k \vec{d}_{nm} \cdot \vec{\Omega}^i) \label{e:hnm}
\end{eqnarray}
In \eqref{hnm}, $\Omega$ can be thought of as both as a solid angle and as a unit vector.

Finally, we observe that carrying both $n$ and $m$ is a bit superfluous, and we'll enumerate the baselines $nm$ 
simply as $j$\footnote{However, we'll need the $nm$ form to deal with closure phase and amplitude later.  It is 
unfortunate that the integer subscript $j$ is overloaded with the imaginary unit.  However, since there are no 
complex quantities in MEM imaging the problem should be minor, and the context clear.}.
So, finally we write
\begin{eqnarray}
V(k\vec{d}_j) &=& B^i h^i_j \\
B^i       &=& B'(\Omega^i) \; d\Omega \\
h^i_j    &=&\exp(j k \vec{d}_j \cdot \vec{\Omega}^i)
\end{eqnarray}
HC6 refer to $h^i_j$ as the point spread function; it's important to note that this is the point spread function for
isotropic antennas only (HC6 later include the antenna gain patterns $A(\Omega)$.)

Measurements:
\begin{eqnarray}
g_j + e_j &=& f^i h^i_j \\
g_j         &=& \textrm{(imperfectly) measured visibility; the data} \\
e_j         &=& \textrm{error in true visibility measurement} \\
f^i           &=& \textrm{modeled image brightness per pointing direction, } \\
h^i_j      &=& \textrm{interferometric point spread function} \\
f^i h^i_j &=& \int_{4\pi} 
\underbrace{ \exp(j k \vec{d_j}\cdot \vec{\Omega}) }_{h^i_j}\; \underbrace{B'(\Omega) \; d\Omega}_{f^i} 
\end{eqnarray}
The last is a restatement of \eqref{vis0}.

Also note that notationally it is not possible for $h^i_j$ to remain complex valued; the real and imaginary 
parts have to be handled explicitly and separately in the Maximum Entropy formulation.  So, if there are 
$N$ antennas, then there are $N(N-1)/2$ baselines, and $N(N-1)$ real and imaginary components.  Thus,
algorithmically, $j$ runs from $0$ to $N(N-1) -1$.
\begin{eqnarray}
h_j^i &=& \cos(k\vec{d}_j \cdot \vec{\Omega}^i ) \;\;\;\;\; \textrm{for } j \textrm{ even} \\ 
h_j^i &=& \sin(k\vec{d}_j \cdot \vec{\Omega}^i) \;\;\;\;\; \textrm{for } j \textrm{ odd}  
\end{eqnarray}
It may be convenient to duplicate each baseline, algorithmically, so that $\vec{d}_{2n} = \vec{d}_{2n+1}$.

The total brightness is assumed to be a measured quantity; it occurs as a self correlation measurement rather
than a correlation between two antennas:
\begin{eqnarray}
F   &=& I^i f^i \\
     &\sim&  \int_{4\pi} B'(\Omega) \; d\Omega
\end{eqnarray}

Below we shall see an additional parameter, $Z$, which provides a normalization and Gibbs function for the MEM 
image.  A close reading of HC6 reveals what appears to be a tautology regarding $F$ and $Z$.  This tautology 
can be avoided by extending the interferometric baselines to include the $0$ baseline, in which the total power of 
the image scene arises naturally within the MEM model.  As an additional benefit, the zero 
baseline demonstrates its use a a "sharpness" or "focussing" parameter.   For small values of $\lambda_j$ ($
\lambda_j \le 1$) the MEM model is nearly linear in the Fourier sense, however for larger values of $\lambda_j$ 
the model can produce very sharp features.  This is one of the strengths of the MEM approach; it can rise to the 
occasion of super resolution given appropriate data.

Image entropy:
\begin{eqnarray}
S &=& -f^i \log(f^i/F) \\
&\sim& \int_{4\pi} B'(\Omega) \log( B'(\Omega)/F) \; d\Omega
\end{eqnarray}

The MaxEnt image model is
\begin{eqnarray}
f^i &=&\frac{F}{Z}\exp(-\lambda_j h_j^i) \\
Z &=& I^i \exp(-\lambda_j h_j^i)
\end{eqnarray}
for some $\lambda_j$ to be determined.  Note that
\begin{equation}
\log(f^i) = -\lambda_j h_j^i
\end{equation}
and we can write the entropy as
\begin{eqnarray}
S &=& -f^i\log((F/Z)\exp(-\lambda_j h_j^i)/F) \\
 &=& f^i \lambda_j h_j^i + f^i I^i\log(Z) \\
 &=& f^i \lambda_j h_j^i + F \log Z
\end{eqnarray}


Functional to be maximized including noise, 

\begin{eqnarray}
E(f(e_j, \lambda_j, \Lambda)) &=& S + \lambda_j(g_j + e_j - f^i h^i_j) + \Lambda(e_j^2 \sigma_j^{-2} - \Sigma) \\
&=& -f^i \log(f^i/F)+ \lambda_j(g_j + e_j - f^i h^i_j) + \Lambda(e_j^2 \sigma_j^{-2} - \Sigma) \nonumber \\
&=& f^i \lambda_j h_j^i + F \log(Z) + \lambda_j(g_j + e_j - f^i h^i_j) + \Lambda(e_j^2 \sigma_j^{-2} - \Sigma) \nonumber \\
&=& \lambda_n(g_j + e_j) + F \log Z + \Lambda(e_j^2 \sigma_j^{-2} - \Sigma) 
\end{eqnarray} 

Differentiating with respect to $\lambda_j, e_j, \Lambda$ we receive these equations to solve:
\begin{eqnarray}
g_j + e_j - f^i h^i_j &=& 0 \\[1em]
\lambda_j + \frac{2\Lambda}{\sigma_j^2} e_j &=& 0 \\[0.5em]
\Lambda^2 - \frac{\lambda_n^2 \sigma_n^2}{4\Sigma} &=& 0
\end{eqnarray}

In the case of no noise, we have
\begin{eqnarray}
g_j - f^i h^i_j &=& 0 
\end{eqnarray}

This could be useful for generating initial conditions.

\section{another whack}

The formulation of HC6 accommodates correlated errors among the $g_j$ through a similarity transform, and 
that's a good thing.

One of the surprising things about he solution of the coupled equations E.35-37 is that, given a set of $g_j$, and 
a solution $\lambda_j$ to them, the numerical solution $\lambda_j$ is extremely stable to variations of $g_j$.  
While initially satisfying, this leads to the puzzling circumstance that very few solutions are actually possible; or, to 
put it differently, in the noiseless case (perfect data $g_j$) the synthesized image is not a smooth function of the 
data.  A little investigation of this reveals that the MaxEnt image model is very stiff.  

\begin{eqnarray}
\Phi &=& \int \exp(-\lambda_k f_k(\Omega)) \;d\Omega \\
\Phi_n &=& \partial \Phi/\partial \lambda_n = \int -f_n(\Omega) \exp(-\lambda_k f_k(\Omega)) \;d\Omega = g_n \\
\Phi_{nm} &=& \partial^2 \Phi/\partial \lambda_n \partial \lambda_m = \int f_n(\Omega) f_m(\Omega) 
\exp(-\lambda_k f_k(\Omega)) \;d\Omega \\
\Phi_{nm} &=& \textrm{symmetric, positive definite} \nonumber \\
\Phi_{nm} \Delta \lambda_m &=& \Delta g_n \\
\Delta \lambda_m &=& [\Phi_{nm}]^{-1} \Delta g_n
\end{eqnarray}

The issue is that $\Phi_{nm}$ (and its inverse) is typically quite stiff --- the condition number is large, the range of
singular values/eigenvalues is large.  The error model is thus not so much estimating actual errors as it is 
compensating for the stiffness of $\Phi_{nm}$.  It would be interesting to find an alternative model which was less 
stiff, which would nevertheless preserve the friendly algebraic properties of the MaxEnt image model.

One possibility is the following:
\begin{eqnarray}
\Phi &=& \int \exp(-\lambda_k f_k(\Omega)) \;d\Omega + \int \exp(-\alpha \lambda_k f_k(\Omega)) \;d\Omega \\
&=&  \int \exp(-\lambda_k f_k(\Omega))\ + \exp(-\alpha \lambda_k f_k(\Omega)) \;d\Omega \\
\Phi_n &=& -\int f_n(\Omega) \exp(-\lambda_k f_k(\Omega)) \;d\Omega + \cdots \nonumber \\
 & & - \; \alpha \int f_n(\Omega) \exp(-\alpha \lambda_k f_k(\Omega)) \;d\Omega \\
\Phi_n &=& -\int f_n(\Omega) [ \exp(-\lambda_k f_k(\Omega)) + \alpha  \exp(-\alpha\lambda_k f_k(\Omega)) ] \;d
\Omega \\
\Phi_{nm} &=& \int f_n(\Omega) f_m(\Omega) \exp(-\lambda_k f_k(\Omega)) \;d\Omega + \cdots \nonumber \\
 & & + \; \alpha^2 \int f_n(\Omega) f_m(\Omega) \exp(-\alpha \lambda_k f_k(\Omega)) \;d\Omega \\
\Phi_{nm} &=& \!\int \! f_n(\Omega) f_m(\Omega) [ \exp(-\lambda_k f_k(\Omega)) + \alpha^2  \exp(-\alpha
\lambda_k 
f_k(\Omega)) ] \;d \Omega 
\end{eqnarray}
This is clearly extensible.  One possibility would be
\begin{equation}
\Phi = \int e^{-\lambda_k f_k(\Omega)} +  e^{-\alpha \lambda_k f_k(\Omega)} + 
e^{-\beta \lambda_k f_k(\Omega)} + e^{-\gamma \lambda_k f_k(\Omega)} \; d\Omega
\end{equation} 
\printindex

\cleardoublepage

\phantomsection

\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{radar}
\bibliographystyle{plain}


\end{document}
